Markdown to estimate the abundance of brain cell types with the reprocessed counts from MSSM cohorts (with TMM-VOOM normalisation)

### 2017-02-03 code/Rmd/MSSM_RNASeq_Reprocessed_TMM_VOOM_CellProp.Rmd
Added markdown to deconvolute celltype fraction from MSSM data

---
title: "Markdown to estimate the abundance of brain cell types with the reprocessed counts from Mayo cohorts (with TMM-VOOM normalisation)"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(synapseClient)
library(knit2synapse)

@@

knit2synapse::knitToFolderEntity(file = "./Mayo_RNASeq_Reprocessed_TMM_VOOM_CellProp.Rmd",
                                 parentId = "syn5570325",
                                 entityName = 'MAYO Cell Type Abundance with Reprocessed RNASeq')
```

```{r libs, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
## It is assumed your working directory is where this file

@@

```{r synapse.parameters, include=FALSE, cache=TRUE}
# Synapse parameters
parentId = 'syn5570325';
activityName = 'Cell type abundance in Mayo';
activityDescription = 'Cell type abundance in Mayo with TMM-VOOM normalised reprocessed RNASeq counts';

thisFileName <- 'Mayo_RNASeq_Reprocessed_TMM_VOOM_CellProp.Rmd'

# Github link
thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='celltype')
thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('code/Rmd/',thisFileName))
```
### Data download
#### Obtain count matrix and metadata from synapse.
```{r download.data, cache=TRUE}
# Get reprocessed sample counts for temporal cortex
COUNT_TC_ID <- 'syn7987765';
ALL_USED_IDs <- COUNT_TC_ID
COUNT_TC <- read.table(synGet(COUNT_TC_ID)@filePath, header=T, sep='\t', check.names = F)
colnames(COUNT_TC) = paste(colnames(COUNT_TC), 'TCX', sep = '_')

# Get reprocessed sample counts for cerebellum
COUNT_CE_ID <- 'syn7513807';
ALL_USED_IDs[length(ALL_USED_IDs)+1] <- COUNT_CE_ID
COUNT_CE <- read.table(synGet(COUNT_CE_ID)@filePath, header=T, sep='\t', check.names = F)

COUNT = cbind(COUNT_TC, COUNT_CE[rownames(COUNT_TC),])

# Get clinical metadata
METADATA_TC_ID <- 'syn3817650'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_TC_ID
METADATA_TC <- read.table(synGet(METADATA_TC_ID)@filePath,sep='\t',header=T, stringsAsFactors=F) %>%
  tidyr::separate(ID, c('Donor_ID', 'BrainRegion'), sep = '_') %>%
  dplyr::mutate(ID = paste(Donor_ID, BrainRegion, sep = '_'))

# Get clinical metadata
METADATA_CE_ID <- 'syn5223705'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_CE_ID
METADATA_CE <- read.table(synGet(METADATA_CE_ID)@filePath,sep='\t',header=T, stringsAsFactors=F) %>%
  tidyr::separate(SampleID, c('Donor_ID', 'BrainRegion')) %>%
  dplyr::mutate(ID = paste(Donor_ID, BrainRegion, sep = '_')) %>%
  dplyr::rename(Gender = Sex, FLOWCELL = Flowcell)

METADATA = rbindlist(list(METADATA_TC, METADATA_CE), use.names = T, fill = T) %>%
  as.data.frame
```

### Data preprocessing
```{r preprocess.data,cache=FALSE, echo=TRUE}
# Fix metadata
METADATA$AgeAtDeath = gsub("_or_above", "", METADATA$AgeAtDeath)
METADATA$Diagnosis = gsub("Pathologic Aging","Pathologic_Aging", METADATA$Diagnosis)
METADATA$BrainRegion.Diagnosis = paste(METADATA$BrainRegion, METADATA$Diagnosis, sep = '.')
METADATA$RIN2 = METADATA$RIN^2
METADATA <- METADATA %>%
  filter(!is.na(Source)) %>%
  dplyr::filter(!(ID %in% c("11311_CER", "1923_CER", "1923_TCX", 
                            "11396_TCX", "11294_TCX", "11408_TCX")))
```
Following samples were considered outliers and removed based on PCA and/or eucledian distance: 11311\_CER, 1923\_CER, 1923\_TCX, 11396\_TCX, 11294\_TCX, 11408\_TCX
```{r preprocess.data1}
# Match covariates to expression data
indToRetain = intersect(METADATA$ID, colnames(COUNT))
indRemoved = setdiff(colnames(COUNT), METADATA$ID)

@@

rownames(METADATA) = METADATA$ID
METADATA = METADATA[indToRetain,]
```
`r dim(COUNT)[2]` samples from `r length(unique(METADATA$Donor_ID))` subjects were obtained from the Mayo cohorts in AMP-AD reprocessed rnaseq project.

Following samples were removed `r indRemoved`

### Covariate clustering
Determine relationship between covariates. 
```{r covariates.clustering}
FactorCovariates <- c("Donor_ID", "Source", "Gender", "FLOWCELL", "BrainRegion.Diagnosis")
ContCovariates <- c("RIN", "RIN2", "AgeAtDeath","PMI")

# Find inter relation between factor covariates
COVARIATES = METADATA[,c(FactorCovariates,ContCovariates),drop=F]
rownames(COVARIATES) <- METADATA$ID

@@

Remove genes that have less than 1 cpm counts in at least 50% of samples per BrainRegion.Diagnosis
```{r cpmnormalisation}
genesToAnalyze = COVARIATES %>%
  rownameToFirstColumn('Sampleid') %>%
  dlply(.(BrainRegion.Diagnosis), .fun = function(mtd, count){
    processed.counts = getGeneFilteredGeneExprMatrix(count[,mtd$Sampleid],
                                                     MIN_GENE_CPM=1, 
                                                     MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0.5)
    processed.counts$filteredExprMatrix$genes
    }, COUNT)
genesToAnalyze = unlist(genesToAnalyze) %>% unique()

@@

```{r get.gene.symbols, cache=TRUE, include=FALSE}
backgroundGenes = data.frame(gene.id = rownames(VOOM.GENE_EXPRESSION$E)) %>%
  dplyr::mutate(ensembl_gene_id = gene.id) %>%
  tidyr::separate(ensembl_gene_id, c('ensembl_gene_id', 'position'), sep = '\\.')

@@

```{r plot.cellprop}
cellprop1 = as.data.frame(cellprop)
cellprop1$cogdx = COVARIATES[rownames(cellprop), 'cogdx']
cellprop1 = cellprop1 %>% 
  rownameToFirstColumn('SampleID') %>%
  dplyr::select(-`P-value`,-Correlation, -RMSE) %>%
  tidyr::gather(Celltype, Fraction, -SampleID, -cogdx) %>%
  dplyr::arrange(desc(cogdx))

p = ggplot(cellprop1, aes(x = SampleID, y = Fraction, fill = Celltype, color = cogdx)) + geom_bar(stat = 'identity')
p = p + theme(axis.text.x = element_text(angle = 90, hjust = 1))
# p = p + facet_grid(.~cogdx)
# p

p = ggplot(cellprop1, aes(x = Celltype, y = Fraction, color = cogdx)) + geom_boxplot()
p = p + theme(axis.text.x = element_text(angle = 90, hjust = 1))
p
```
### Synapse store
```{r syn.store, cache=TRUE}
CODE = Folder(name = 'MAYO Cell Type Abundance with Reprocessed RNASeq', parentId = parentId)
CODE = synStore(CODE)

cellprop %>%
  rownameToFirstColumn('SampleID') %>%
  write.table(file = 'MAYO_CellFraction_Reproc.tsv', row.names = F, sep = '\t', quote = F)
obj = File('MAYO_CellFraction_Reproc.tsv', name = 'MAYO Cell Fraction', parentId = CODE$properties$id)
obj = synStore(obj, activityName = activityName, activityDescription = activityDescription,
               used = ALL_USED_IDs, executed = thisFile)
```

>>>>

---
title: "Markdown to estimate the abundance of brain cell types with the reprocessed counts from MSSM cohorts (with TMM-VOOM normalisation)"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(synapseClient)
library(knit2synapse)

@@

knit2synapse::knitToFolderEntity(file = "./MSSM_RNASeq_Reprocessed_TMM_VOOM_CellProp.Rmd",
                                 parentId = "syn5570325",
                                 entityName = 'MSSM Cell Type Abundance with Reprocessed RNASeq')
```

```{r libs, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE, cache=FALSE}
## It is assumed your working directory is where this file

@@

```{r synapse.parameters, include=FALSE, cache=TRUE}
# Synapse parameters
parentId = 'syn5570325';
activityName = 'Cell type abundance in MSSM';
activityDescription = 'Cell type abundance in MSSM with TMM-VOOM normalised reprocessed RNASeq counts';

thisFileName <- 'MSSM_RNASeq_Reprocessed_TMM_VOOM_CellProp.Rmd'

# Github link
thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='celltype')
thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('code/Rmd/',thisFileName))
```
### Data download
Data were obtained from AMP-AD portal of MSSM, and following brain regions were analysed for differential expression between case and control:
  | Region | Name                           | 
  | :----- | ------------------------------ | 
  | BM_10  | Frontal Pole (FP)              |
  | BM_22  | Superior Temporal Gyrus (STG)  |
  | BM_36  | Parahipocampal Gyrus (PHG)     |
  | BM_44  | inferior frontal Gyrus (IFG)   |

* BM 10: frontopolar prefrontal cortex, rostrolateral prefrontal cortex, or anterior prefrontal cortex, 
* BM 22: In humans it corresponds approximately to the lateral and caudal two thirds of the superior temporal gyrus, 
* BM 36: Ectorhinal area 36 is a subdivision of the cytoarchitecturally defined temporal region of cerebral cortex,
* BM 44: pars opercularis (of the inferior frontal gyrus), and it refers to a subdivision of the frontal region of cerebral cortex.

#### Obtain count matrix and metadata from synapse.
```{r download.data, cache=TRUE}
# Get reprocessed sample counts for temporal cortex
COUNT_ID <- 'syn7994853';
ALL_USED_IDs <- COUNT_ID
COUNT <- read.table(synGet(COUNT_ID)@filePath, header=T, sep='\t', check.names = F)

# Get sample ids
SampleID = data.frame(SampleID = colnames(COUNT),
                 ID = colnames(COUNT)) %>% 
  tidyr::separate(ID, c('A','B','ID'), sep = '_')

SampleID = SampleID %>%
  dplyr::select(SampleID, ID) %>%
  dplyr::mutate(ID = as.numeric(ID))

# Get technical metadata
METADATA_TECH_ID <- 'syn6100548'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_TECH_ID
METADATA_TECH <- read.table(synGet(METADATA_TECH_ID)@filePath,sep=',',header=T, row.names=1) %>%
  group_by(sampleIdentifier) %>%
  top_n(1, -rRNA.rate) %>%
  dplyr::select(sampleIdentifier, BrodmannArea, barcode, individualIdentifier, batch, RIN, TotalReads, Mapped, rRNA.rate) %>%
  unique

# Get clinical metadata
METADATA_CLINICAL_ID <- 'syn6101474'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_CLINICAL_ID
METADATA_CLINICAL <- read.csv(synGet(METADATA_CLINICAL_ID)@filePath, header = T)

METADATA <- full_join(METADATA_TECH, METADATA_CLINICAL)
```

### Data preprocessing
```{r preprocess.data,cache=FALSE, echo=TRUE}
# Fix metadata
METADATA$AOD = gsub('\\+','',METADATA$AOD)
METADATA$Dx[METADATA$CDR <= 1] = 'ND'
METADATA$Dx[METADATA$CDR == 2 | METADATA$CDR == 3] = 'MD'
METADATA$Dx[METADATA$CDR >= 4] = 'SD'
METADATA$BrodmannArea.Dx = paste(METADATA$BrodmannArea, METADATA$Dx, sep = '.')
METADATA$RIN2 = METADATA$RIN^2
levels(METADATA$batch)[levels(METADATA$batch) == ''] = 'NoBatch'

METADATA <- METADATA %>%
  filter(!is.na(BrodmannArea), !is.na(PMI), !is.na(RIN))
```

```{r preprocess.data1}
# Match covariates to expression data
indToRetain = intersect(METADATA$sampleIdentifier, colnames(COUNT))
indRemoved = setdiff(colnames(COUNT), METADATA$sampleIdentifier)

@@

METADATA = as.data.frame(METADATA)
rownames(METADATA) = METADATA$sampleIdentifier
METADATA = METADATA[indToRetain,]
```
`r dim(COUNT)[2]` samples from `r length(unique(METADATA$individualIdentifier))` subjects were obtained from the MSBB cohorts in AMP-AD reprocessed rnaseq project

Following `r length(indRemoved)` samples are removed:`r paste(indRemoved, collapse = ', ')`
### Covariate clustering
Determine relationship between covariates. 
```{r covariates.clustering}
FactorCovariates <- c("individualIdentifier", "batch", "RACE", "SEX", "BrodmannArea.Dx")
ContCovariates <- c("RIN", "RIN2", "AOD","PMI", "TotalReads", "Mapped", "rRNA.rate")

# Find inter relation between factor covariates
COVARIATES = METADATA[,c(FactorCovariates,ContCovariates),drop=F]
rownames(COVARIATES) <- METADATA$sampleIdentifier

@@

Remove genes that have less than 1 cpm counts in at least 50% of samples per BrodmannArea.Dx
```{r cpmnormalisation}
genesToAnalyze = COVARIATES %>%
  rownameToFirstColumn('sampleIdentifier') %>%
  dlply(.(BrodmannArea.Dx), .fun = function(mtd, count){
    processed.counts = getGeneFilteredGeneExprMatrix(count[,mtd$sampleIdentifier],
                                                     MIN_GENE_CPM=1, 
                                                     MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0.5)
    processed.counts$filteredExprMatrix$genes
    }, COUNT)
genesToAnalyze = unlist(genesToAnalyze) %>% unique()

@@

```{r get.gene.symbols, include=FALSE}
backgroundGenes = data.frame(gene.id = rownames(VOOM.GENE_EXPRESSION$E)) %>%
  dplyr::mutate(ensembl_gene_id = gene.id) %>%
  tidyr::separate(ensembl_gene_id, c('ensembl_gene_id', 'position'), sep = '\\.')

@@

```{r plot.cellprop, fig.width=16, fig.height=8}
cellprop1 = as.data.frame(cellprop)
cellprop1$BrodmannArea.Dx = COVARIATES[rownames(cellprop), 'BrodmannArea.Dx']
cellprop1 = cellprop1 %>% 
  rownameToFirstColumn('SampleID') %>%
  dplyr::select(-`P-value`,-Correlation, -RMSE) %>%
  tidyr::gather(Celltype, Fraction, -SampleID, -BrodmannArea.Dx) %>%
  dplyr::arrange(desc(BrodmannArea.Dx))

p = ggplot(cellprop1, aes(x = SampleID, y = Fraction, fill = Celltype, color = BrodmannArea.Dx)) + geom_bar(stat = 'identity')
p = p + theme(axis.text.x = element_text(angle = 90, hjust = 1))
# p = p + facet_grid(.~BrodmannArea.Dx)
# p

p = ggplot(cellprop1, aes(x = Celltype, y = Fraction, color = BrodmannArea.Dx)) + geom_boxplot()
p = p + theme(axis.text.x = element_text(angle = 90, hjust = 1))
p
```
### Synapse store
```{r syn.store, cache=TRUE}
CODE = Folder(name = 'MSSM Cell Type Abundance with Reprocessed RNASeq', parentId = parentId)
CODE = synStore(CODE)

cellprop %>%
  rownameToFirstColumn('SampleID') %>%
  write.table(file = 'MSSM_CellFraction_Reproc.tsv', row.names = F, sep = '\t', quote = F)
obj = File('MSSM_CellFraction_Reproc.tsv', name = 'MSSM Cell Fraction', parentId = CODE$properties$id)
obj = synStore(obj, activityName = activityName, activityDescription = activityDescription,
               used = ALL_USED_IDs, executed = thisFile)
```
